The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-08-07 paper.tex
more paper

The Terminologia anatomica~\citep{ta2} (TA) is a hierarchy of anatomical structure concepts for the entire human body.
For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
In some cases, Latin and English synonyms are also included, e.g. \emph{malar bone} is an English synonym of the Latin preferred term \emph{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
For instance, the \emph{processus zygomaticus} can be found both at the \emph{os frontale} and the \emph{os temporale}.
Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
For example, \emph{ossa digitorum manus / pedis} (engl. \emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.

The \emph{Foundational Model of Anatomy} (FMA)~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.

@@

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo})

@@

% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNOdc ontology}\label{fig:core}
\end{figure}

\begin{table}
  \centering
  \caption{Classes of ANNOdc along with interlinks to the FMA and the GFO top-level ontology.
  \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
  \label{tab:core}
  \begin{tabulary}{\textwidth}{llLl}
    \toprule
	ANNOdc class					&ANNOds Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&\anno{Skeleton}		&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomicalEntity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{i}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{MeasurementPoint}			&\anno{Lambda}\\%, \anno{Ophistion}\\
	\anno{OrientationPoint}			&\anno{Orbitale}\\
	\anno{AnatomicalPlane}			&\anno{FrontalPlane}	&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 	&\anno{SagittalAxis}\\
	%\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&\anno{Mandibula}				&Bone organ \fma{5018}\\
	\anno{BonePart}					&\anno{ArcusAlveolaris}	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&\anno{Cranium}				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}\\%			&\anno{DentesPermanentes}\\%		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}					&\anno{DensCaninus}			&Tooth \fma{12516}\\
	\anno{ToothPart}\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}				&\anno{SexGiles19671}			&											&\gfo{Attributive}\\
	\anno{AnatomicalProperty}		&						&													&\gfo{Attributive}\\
	%\anno{AnatomicalProperty}		&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&\anno{Sinister}	&Anatomical qualitative coordinate \fma{30346}	&\gfo{Relator}\\
	%\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
% yes, walker method is published in buikstra
Method after Walker~\cite{datacollection}.
Image based on \cite{datacollection}.
}\label{fig:gfo}
\end{figure}

@@

\section{Development of ANNOds}\label{sec:domain}
While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.

@@

@@

General considerations 
All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.

@@

Process for creating definitions and measurements (Headline to be deleted) 
For the definitions and measurements, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or ancient Greek terms were missing from the English-language literature.
The FMA also contains only English terms.
For this reason, these were also searched out with their synonyms.
Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
This required a positional description (e.g. Punctum superioris capitis femoris as the superior located point of the Caput femoris) necessary.
The examination of the Latin grammar was relevant for this.
For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.

%\subsection{Functions} Headline to be deleted
The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.

@@

%Sex determination Headline to be deleted 
% auto translated from the project report, use as basis:
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.

Regress functions Headline to be deleted 
Regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.

>>>>

\subsection{Terminologia anatomica (TA)}
The TA~\citep{ta2} is a hierarchy of anatomical structure concepts for the entire human body.
For historical reasons, it uses a terminology consisting of Latin and originally Greek, which was later latinised.
For each anatomical structure, the TA provides the \enquote{preferred}, i.e. standard, Latin term and its English equivalent as well as an individual identification number.
In some cases, Latin and English synonyms are also included, e.g. \emph{malar bone} is an English synonym of the Latin preferred term \emph{Os zygomaticum} and its English equivalent \emph{zygomatic bone}.
The identification number is crucial, as certain landmarks are mostly listed without bone affiliation, resulting in multiple occurrences. %(s. Bone Part)
For instance, the \emph{processus zygomaticus} can be found both at the \emph{os frontale} and the \emph{os temporale}.
Eponyms are terms derived from proper nouns such as the name of a person, place or thing, and they are considered a type of synonyms since they stand for the same anatomical structure.
For example, \emph{ossa digitorum manus / pedis} (engl. \emph{phalanges of the hand / foot}) are the small bones of the fingers and toes and are also denoted as \emph{phalanges manus / pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
In addition to the outdated print edition~\citep{ta1998}, an extended second edition (TA2) is available online~\citep{ta2}.
Furthermore, there is a commercially available independent companion print publication with German~\citep{anatomylexicon}, English~\citep{pocketatlas} and Spanish~\citep{taspanish} editions of which at least the German edition is regularly updated.
This book contains textual and visual descriptions, whereby anatomical structures are indicated through lines on the illustrations.

\subsection{Foundational Model of Anatomy (FMA)}
The FMA~\citep{fma} ontology represents the physical organisation of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganising the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.

@@

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},

@@

% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{The ANNOdc domain-core ontology.}\label{fig:core}
\end{figure}

\begin{table}
  \centering
  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.
  \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
  \label{tab:core}
  \begin{tabulary}{\textwidth}{llLl}
    \toprule
	ANNOdc class					&ANNOds Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&\anno{Skeleton}		&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomicalEntity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{i}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{MeasurementPoint}			&\anno{Lambda}\\%, \anno{Ophistion}\\
	\anno{OrientationPoint}			&\anno{Orbitale}\\
	\anno{AnatomicalPlane}			&\anno{FrontalPlane}	&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 	&\anno{SagittalAxis}\\
	%\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&\anno{Mandibula}				&Bone organ \fma{5018}\\
	\anno{BonePart}					&\anno{ArcusAlveolaris}	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&\anno{Cranium}				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}\\%			&\anno{DentesPermanentes}\\%		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}					&\anno{DensCaninus}			&Tooth \fma{12516}\\
	\anno{ToothPart}\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}				&\anno{SexGiles19671}			&											&\gfo{Attributive}\\
	\anno{AnatomicalProperty}		&						&													&\gfo{Attributive}\\
	%\anno{AnatomicalProperty}		&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&\anno{Sinister}	&Anatomical qualitative coordinate \fma{30346}	&\gfo{Relator}\\
	%\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
% yes, walker method is published in buikstra
Method after Walker~\cite{datacollection}.
Image based on \cite{datacollection}.
}\label{fig:phenotype}
\end{figure}

@@

\section{Development of ANNOds}\label{sec:domain}

@@

While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.

@@

%\subsections{General considerations}
All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.

@@

\subsection{Process for creating definitions and measurements}
For the definitions and measurements, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or ancient Greek terms were missing from the English-language literature.
The FMA also contains only English terms.
For this reason, these were also searched out with their synonyms.
Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
This required a positional description (e.g. Punctum superioris capitis femoris as the superior located point of the Caput femoris) necessary.
The examination of the Latin grammar was relevant for this.
For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.

\subsection{Functions}
The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.

@@

\subsection{Sex determination}
% auto translated from the project report, use as basis:
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.

\subsection{Regress functions}
Regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnic group or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
